Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-08-06 Report/Sections/simulation.tex
Update Report

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}. The ground-truth parameters differ slightly from those learned in the case study primarily for simplicity of presentation. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $Z^{*(2)}$ is set as a scalar for additional simplicity. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model. This allows us to judge how accurate each model is when estimating its own uncertainty.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
    %
    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
    %
    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
    %
    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
\end{gather*}
%
Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.

The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2. The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2. This corresponds to much more vigorous and variable periodic behavior in the acceleration data for subdive state 2.

@@

As expected, the full CarHHMM-DFT is the best-performing model. Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations. For dive type the average confidence was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$. For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation. The estimated standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3). This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.

>>>>

To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT. The parameters used to generate the data are loosely based on those estimated in the case study in Section \ref{sec:case_study} (see table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made to simplify presentation. For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives. This allows meaningful conclusions to be made for both dive types. In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set as a scalar for additional simplicity. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times. To assess how accurate each model is when estimating its own uncertainty, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.

@@

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
$$
    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
$$
%
After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
%
The observations $Z^*$ are generated using the following parameters:
%
\begin{gather*}
    \mu_A^{*(1)} = 0.0 s, \enspace \sigma_A^{*(1)} = 0.05s, \enspace \phi_A^{*(1)} = 0.99, \\
    %
    \mu_A^{*(2)} = 0.0 s, \enspace \sigma_A^{*(2)} = 0.1s, \enspace \phi_A^{*(2)} = 0.95, \\
    %
    \mu_W^{*(1)} = 10.10, \quad \sigma_W^{*(1)} = 3.18, \\
    %
    \mu_W^{*(2)} = 305.94, \quad \sigma_W^{*(2)} = 17.49.
\end{gather*}
%
Recall that $\Zone$ follows a CarHMM and is Normally distributed while $\Ztwo$ follows an HMM and is Gamma distributed. It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.

The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2. The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1. This corresponds to much more vigorous and variable periodic behavior in the acceleration data.

@@

As expected, since it is the generating model, the full CarHHMM-DFT is the best-performing model. Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations. For dive type the average confidence was above 0.9 (see Table \ref{table:accuracy}). All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased. Even still, these biases are the least severe for the CarHHMM-DFT among the four models. The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error. See Section S-1.2, tables 2-5 for more information.

The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_A^*$. For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation. The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3). This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.
